### 2024-11-27 parts/correction.tex
better knowledges

\AP To address our first goal, let us define that a function $q$ is a
\intro{pumping pattern} from $\Nat^p$ to $\Sigma^*$
whenever there exists words $\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words
$u_1, \dots, u_p \in \Sigma^*$, such that $q(X_1, \dots, X_p) = \alpha_0
\prod_{i = 1}^p u_i^{X_i} \alpha_i$. That is, $q$ is syntactically defined by a
non-commutative monomial over the monoid $\Sigma^*$. \kl{Polynomial functions}
are \kl{commutative} \kl{polyregular functions}.

@@

\begin{lemma}
	\label{binomial-function-star-free:lem}
	Let $F$ be a \kl{binomial function} from $\Nat^k$ to $\Nat$.
	Then it is \kl{represented} by a \kl{star-free polyregular function}.
\end{lemma}
\begin{proof}
	Because of the stability properties of
	$\NSF$ (\cref{stability-polyregular:lemma}), we only need to
	check that given $r,s \in \Nat$,
	the function $x \mapsto \binom{x - r}{s}$ is \kl{represented}
	by a function $f_{r,s} \in \NSF$.
	Let us prove it when $r = 0$, since the other functions
	can be obtained by translating $f_{r,s}$.
    By definition, $\binom{x}{s} = \card{P_{x,s}}$,
    where $P_{x,s} \defined \setof{ (x_1, \dots, x_s) \in \Nat^s }{ 1 \leq x_1 < \cdots < x_s \leq x }$.
    Let us proceed as in \cref{q-polynomial-n-poly:ex} and define 
    $D_{w,s} \defined \setof{ (u_1, \dots, u_s, u_{s+1}) \in (\Sigma^+)^s \times \Sigma^* }{ w = u_1 u_2 \cdots u_s u_{s+1} }$.
    It is clear that $D_{a^x,s}$ is in bijection with $P_{x,s}$ for all $x \in \Nat$
    using the map $(x_1, \dots, x_s) \mapsto (a^{x_1}, \dots, a^{x_s})$.
    Now, using the monoid $M \defined (\set{0,1}, \max)$
    and the morphism $\mu(\varepsilon) \defined 0$ and $\mu(a) \defined 1$,
    one can compute $\card{D_{w,s}}$ 
    as $\pi^\dagger (w)$ where $\pi \colon M^{s+1} \to \Nat$ is defined by
    $\pi(m_1, \dots, m_s, m_{s+1}) \defined m_1 \times \cdots \times m_s$.
    We conclude that $f_{0,s} \in \NSF[s]$ is a \kl{star-free polyregular function}.
\end{proof}

@@

\begin{corollary}
	\label{n-rat-correct:lem}
	Let $P \in \Rel[X_1, \dots, X_p]$ be an \kl{$\Nat$-rational polynomial}.
	Then,
	$P \in \CorrectPoly$.
\end{corollary}
\begin{proof}
    Let $f$ be a \kl{commutative} \kl{$\Nat$-rational series}
	with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
	that \kl{represents} $P$.
    Because $f$ has \kl{polynomial growth},
    $f \in \NPoly$ 
    (\cref{polyregular-polynomial-growth:lemma}).
	Using \cref{n-poly-combinatorics:lem},
	there exists a number $\omega \in \Nat_{\geq 1}$
	and \kl{natural binomial function} $Q$
	such that
	for all $n_1, \dots, n_p \geq 1$:
	\begin{equation*}
		f\left(
		\prod_{i = 1}^p a_i^{\omega n_i}
		\right)
		= Q(n_1, \dots, n_p)
		= P(\omega n_1, \dots, \omega n_p)
		\quad .
	\end{equation*}
    For large enough values of $X$, the \kl{binomial function}
	$\binom{X - p}{k}$ coincides with a polynomial whose leading coefficient
	is $1/k!$ which is \kl{non-negative}.
	We conclude that
	the \kl{maximal monomials} of
	$P(\omega X_1, \dots, \omega X_p)$ are \kl{non-negative},
	and since $\omega \geq 1$, we conclude that
	the \kl{maximal monomials} of $P$ have \kl{non-negative} coefficients.

>>>>

\AP To address our first goal, let us define that a function $q$ is a
\intro{pumping pattern} from $\Nat^p$ to $\Sigma^*$
whenever there exists words $\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words
$u_1, \dots, u_p \in \Sigma^*$, such that $q(X_1, \dots, X_p) = \alpha_0
\prod_{i = 1}^p u_i^{X_i} \alpha_i$. That is, $q$ is syntactically defined by a
non-commutative monomial over the monoid $\Sigma^*$. \kl{Pumping patterns}
are \kl{commutative} \kl{polyregular functions}.

@@

\begin{lemma}
	\label{binomial-function-star-free:lem}
	Let $F$ be a \kl{simple binomial function} from $\Nat^k$ to $\Nat$.
	Then it is \kl{represented} by a \kl{star-free polyregular function}.
\end{lemma}
\begin{proof}
	Because of the stability properties of
	$\NSF$ (\cref{stability-polyregular:lemma}), we only need to
	check that given $r,s \in \Nat$,
	the function $x \mapsto \binom{x - r}{s}$ is \kl{represented}
	by a function $f_{r,s} \in \NSF$.
	Let us prove it when $r = 0$, since the other functions
	can be obtained by translating $f_{r,s}$.
    By definition, $\binom{x}{s} = \card{P_{x,s}}$,
    where $P_{x,s} \defined \setof{ (x_1, \dots, x_s) \in \Nat^s }{ 1 \leq x_1 < \cdots < x_s \leq x }$.
    Let us proceed as in \cref{q-polynomial-n-poly:ex} and define 
    $D_{w,s} \defined \setof{ (u_1, \dots, u_s, u_{s+1}) \in (\Sigma^+)^s \times \Sigma^* }{ w = u_1 u_2 \cdots u_s u_{s+1} }$.
    It is clear that $D_{a^x,s}$ is in bijection with $P_{x,s}$ for all $x \in \Nat$
    using the map $(x_1, \dots, x_s) \mapsto (a^{x_1}, \dots, a^{x_s})$.
    Now, using the monoid $M \defined (\set{0,1}, \max)$
    and the morphism $\mu(\varepsilon) \defined 0$ and $\mu(a) \defined 1$,
    one can compute $\card{D_{w,s}}$ 
    as $\pi^\dagger (w)$ where $\pi \colon M^{s+1} \to \Nat$ is defined by
    $\pi(m_1, \dots, m_s, m_{s+1}) \defined m_1 \times \cdots \times m_s$.
    We conclude that $f_{0,s} \in \NSF[s]$ is a \kl{star-free polyregular function}.
\end{proof}

@@

\begin{corollary}
	\label{n-rat-correct:lem}
	Let $P \in \Rel[X_1, \dots, X_p]$ be an \kl{$\Nat$-rational polynomial}.
	Then,
	$P \in \CorrectPoly$.
\end{corollary}
\begin{proof}
    Let $f$ be a \kl{commutative} \kl{$\Nat$-rational series}
	with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
	that \kl{represents} $P$.
    Because $f$ has \kl{polynomial growth},
    $f \in \NPoly$ 
    (\cref{polyregular-polynomial-growth:lemma}).
	Using \cref{n-poly-combinatorics:lem},
	there exists a number $\omega \in \Nat_{\geq 1}$
	and \kl{natural binomial function} $Q$
	such that
	for all $n_1, \dots, n_p \geq 1$:
	\begin{equation*}
		f\left(
		\prod_{i = 1}^p a_i^{\omega n_i}
		\right)
		= Q(n_1, \dots, n_p)
		= P(\omega n_1, \dots, \omega n_p)
		\quad .
	\end{equation*}
    For large enough values of $X$, the \kl{simple binomial function}
	$\binom{X - p}{k}$ coincides with a polynomial whose leading coefficient
	is $1/k!$ which is \kl{non-negative}.
	We conclude that
	the \kl{maximal monomials} of
	$P(\omega X_1, \dots, \omega X_p)$ are \kl{non-negative},
	and since $\omega \geq 1$, we conclude that
	the \kl{maximal monomials} of $P$ have \kl{non-negative} coefficients.

### 2024-11-27 parts/counter.tex
better knowledges

\begin{theorem}[{\cite[Theorem 1.7]{BOJA18}}]
    \label{pre-image-regular:fact}
    The pre-image of a regular language by a \kl{polyregular function}
    is a regular language.
\end{theorem}

@@

Please note that the same argument cannot be leveraged for proving that $P$ is
not a \kl{$\Rel$-rational polynomial}: \cref{pre-image-regular:fact}
only holds for
\emph{string-to-string} functions, and is applied to the specific case where
the output alphabet is $\set{1}$, i.e., where the output of the function
belongs to $\set{1}^*$ which is isomorphic to $\Nat$.

>>>>

\begin{theorem}[{\cite[Theorem 1.7]{BOJA18}}]
    \label{pre-image-regular:fact}
    The pre-image of a regular language by a (string-to-string) \kl{polyregular function}
    is a regular language.
\end{theorem}

@@

Please note that the same argument cannot be leveraged for proving that $P$ is
not representd by a \kl{$\Rel$-rational series}: \cref{pre-image-regular:fact}
only holds for
\emph{string-to-string} functions, and is applied to the specific case where
the output alphabet is $\set{1}$, i.e., where the output of the function
belongs to $\set{1}^*$ which is isomorphic to $\Nat$.

### 2024-11-27 parts/newintro.tex
better knowledges

\AP Given a semiring $\mathbb{S}$, and a finite alphabet $\Sigma$, the class of
\kl{(noncommutative) $\mathbb{S}$-rational series} is defined as functions from
$\Sigma^*$ to $\mathbb{S}$ that are computed by \kl{$\mathbb{S}$-weighted
automata} \cite{BERE10}. This computational model is a generalization of the
classical notion of non-deterministic finite automata to the weighted setting,
where transitions are labeled with elements of $\mathbb{S}$. The semantics of
\kl{$\mathbb{S}$-weighted automata} on a given word $w$ is defined by the sum
over all accepting runs reading $w$, of the product of the weights of the
transitions taken along this run. In this paper, we are interested in the case
where $\mathbb{S}$ equals $\Nat$ or $\Rel$, hence, in $\Nat$-rational series
($\intro*\NRat$) and $\Rel$-rational series ($\intro*\ZRat$). It is clear that
$\NRat$ is a proper subclass of $\ZRat$, and a longstanding open problem is to
provide an algorithm that decides whether a given $\ZRat$ is in $\NRat$
\cite{KARH77}.

@@

\Cref{n-in-z-rat:problem} recently received attention in the context of
\kl{polyregular functions}, a computational model that aims to
generalize the theory of regular languages to the setting of string-to-string
functions \cite{BOJA18}. In the case of regular languages, \emph{star-free
languages} form a robust subclass of regular languages described equivalently
in terms of first order logic \cite{MNPA71}, counter-free automata
\cite{MNPA71}, or aperiodic monoids \cite{SCHU65}. Analogously, there exists a
\emph{star-free} fragment of \kl{polyregular functions} called \kl{star-free
polyregular functions} \cite{BOJA18}. One open question in this area is to
decide whether a given \kl{polyregular function} is \kl{star-free}.

>>>>

\AP Given a semiring $\mathbb{S}$, and a finite alphabet $\Sigma$, the class of
\intro{(noncommutative) $\mathbb{S}$-rational series} is defined as functions from
$\Sigma^*$ to $\mathbb{S}$ that are computed by \kl{$\mathbb{S}$-weighted
automata} \cite{BERE10}. This computational model is a generalization of the
classical notion of non-deterministic finite automata to the weighted setting,
where transitions are labeled with elements of $\mathbb{S}$. The semantics of
\kl{$\mathbb{S}$-weighted automata} on a given word $w$ is defined by the sum
over all accepting runs reading $w$, of the product of the weights of the
transitions taken along this run. In this paper, we are interested in the case
where $\mathbb{S}$ equals $\Nat$ or $\Rel$, hence, in $\Nat$-rational series
($\intro*\NRat$) and $\Rel$-rational series ($\intro*\ZRat$). It is clear that
$\NRat$ is a proper subclass of $\ZRat$, and a longstanding open problem is to
provide an algorithm that decides whether a given $\ZRat$ is in $\NRat$
\cite{KARH77}.

@@

\Cref{n-in-z-rat:problem} recently received attention in the context of
\kl{polyregular functions} ($\intro*\Poly$), a computational model that aims to
generalize the theory of regular languages to the setting of string-to-string
functions \cite{BOJA18}. In the case of regular languages, \emph{star-free
languages} form a robust subclass of regular languages described equivalently
in terms of first order logic \cite{MNPA71}, counter-free automata
\cite{MNPA71}, or aperiodic monoids \cite{SCHU65}. Analogously, there exists a
\emph{star-free} fragment of \kl{polyregular functions} called \kl{star-free
polyregular functions} ($\intro*\SF$) \cite{BOJA18}. One open question in this area is to
decide whether a given \kl{polyregular function} is \kl{star-free}.

### 2024-11-27 parts/prelim.tex
better knowledges

\AP We assume that the reader is familiar with the basics of automata theory,
in particular the notions of monoid morphisms, idempotents in monoids, monadic
second-order ($\MSO$) logic and first-order ($\FO$) logic over finite words
(see e.g. \cite{THOM97}). As aperiodicity will be a central notion of this
paper, let us recall that a monoid $M$ is \intro(monoid){aperiodic} whenever
for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$. If the
monoid $M$ is finite, this $n$ can be uniformly chosen for all elements in $M$.

@@

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \kl{polyregular functions} for
these two output semigroups, and we refer the reader to \cite{BOKL19} for the
general definition of \kl{polyregular functions}. 
We chose in this paper to provide a combinatorial description of
polyregular functions with commutative outputs because it will play nicely with
our analysis on polynomials. This description is very similar in shape to the
\emph{finite counting automata} introduced by \cite{SCHU62}.

>>>>

\AP We assume that the reader is familiar with the basics of automata theory,
in particular the notions of monoid morphisms, idempotents in monoids, monadic
second-order ($\intro*\MSO$) logic and first-order ($\intro*\FO$) logic over finite words
(see e.g. \cite{THOM97}). As aperiodicity will be a central notion of this
paper, let us recall that a monoid $M$ is \intro(monoid){aperiodic} whenever
for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$. If the
monoid $M$ is finite, this $n$ can be uniformly chosen for all elements in $M$.

@@

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \intro{polyregular functions}
for these two output semigroups, and we refer the reader to \cite{BOKL19} for
the general definition of \kl{polyregular functions} and their aperiodic
counterpart, the \intro{star-free polyregular functions}. We chose in this
paper to provide a combinatorial description of polyregular functions with
commutative outputs because it will play nicely with our analysis on
polynomials. This description is very similar in shape to the \emph{finite
counting automata} introduced by \cite{SCHU62}.

### 2024-11-27 parts/starfree.tex
better knowledges

\begin{definition}
    \label{k-combinatorial:def}
    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
    such that
    for all
    \kl{polynomial functions} $q \colon \Nat^k \to \Sigma^*$,
    there exists a \kl{strongly natural binomial polynomial} $P$
    satisfying:
    \begin{equation*}
        f \circ q(\omega X_1,\dots, \omega X_k)
        = 
        P
        \quad 
        \text{ over } (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}
\end{definition}

@@

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and $f \in \ZPoly[k]$ be \kl{commutative}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    The implication
    \cref{f-npoly-combi:item}
    $\implies$ \cref{f-combinatorial:item} immediately follows
    from 
    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
    For the converse implication, we use \cref{decompose-polynomial:lem} 
    effectively compute a polynomial representation of $f$. That is, we obtain
    an $\omega_1 \in \Nat$ and \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega_1]^k}$. Now, because $f$ is
    \kl{$(k,\Nat)$-combinatorial}, we also have a number $\omega_2 \in \Nat$
    such that for all \kl{polynomial functions} $q$, there exists a
    \kl{strongly natural binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
    = P$ over $(\Nat_{\geq 1})^k$. By considering $\omega \defined
    \omega_1 \times \omega_2$, we can without loss of generality assume that
    $\omega_1 = \omega_2$ in the rest of this proof.

@@

    Let us therefore consider the \kl{polynomial function} $q \colon \Nat^k \to
    \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. Because $f$ is
    \kl{$(k,\Nat)$-combinatorial} (with parameter $\omega$), there exists a
    \kl{strongly natural binomial polynomial} $P$ such that for all $x_1,
    \dots, x_n \geq 1$, the equality $f \circ q(\omega x_1, \dots, \omega x_k)
    = P(x_1, \dots, x_k)$ holds. In particular, we conclude that over
    $(\Nat_{\geq 1})^{\card{S}}$ the polynomial $P_t$ and $P$ coincide. They
    are therefore equal as multivariate polynomials, and we conclude that $P_t$
    is a \kl{strongly natural binomial polynomial} \kl{$\Nat$-rational
    polynomial}.

@@

\begin{definition}[Ultimately polynomial]
    \label{ultimately-polynomial:def}
    Let $\Sigma$ be a finite alphabet. 
    A function $f \colon \Sigma^* \to \Rel$
    is \intro{ultimately polynomial}
    when there exists $N_0 \in \Nat$ such that
    for all $n \in \Nat$,
    for all \kl{polynomial function} $q \colon \Nat^n \to \Sigma^*$,
    there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
    such that:
    \begin{equation*}
        f \circ p = P
        \quad 
        \text{ over } (\Nat_{\geq N_0})^n
        \quad .
    \end{equation*}
\end{definition}

>>>>

\begin{definition}
    \label{k-combinatorial:def}
    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
    such that
    for all
    \kl{pumping patterns} $q \colon \Nat^k \to \Sigma^*$,
    there exists a \kl{strongly natural binomial polynomial} $P$
    satisfying:
    \begin{equation*}
        f \circ q(\omega X_1,\dots, \omega X_k)
        = 
        P
        \quad 
        \text{ over } (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}
\end{definition}

@@

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and $f \in \ZPoly[k]$ be \kl{commutative}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    The implication
    \cref{f-npoly-combi:item}
    $\implies$ \cref{f-combinatorial:item} immediately follows
    from 
    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
    For the converse implication, we use \cref{decompose-polynomial:lem} 
    effectively compute a polynomial representation of $f$. That is, we obtain
    an $\omega_1 \in \Nat$ and \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega_1]^k}$. Now, because $f$ is
    \kl{$(k,\Nat)$-combinatorial}, we also have a number $\omega_2 \in \Nat$
    such that for all \kl{pumping patterns} $q$, there exists a
    \kl{strongly natural binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
    = P$ over $(\Nat_{\geq 1})^k$. By considering $\omega \defined
    \omega_1 \times \omega_2$, we can without loss of generality assume that
    $\omega_1 = \omega_2$ in the rest of this proof.

@@

    Let us therefore consider the \kl{pumping pattern} $q \colon \Nat^k \to
    \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. Because $f$ is
    \kl{$(k,\Nat)$-combinatorial} (with parameter $\omega$), there exists a
    \kl{strongly natural binomial polynomial} $P$ such that for all $x_1,
    \dots, x_n \geq 1$, the equality $f \circ q(\omega x_1, \dots, \omega x_k)
    = P(x_1, \dots, x_k)$ holds. In particular, we conclude that over
    $(\Nat_{\geq 1})^{\card{S}}$ the polynomial $P_t$ and $P$ coincide. They
    are therefore equal as multivariate polynomials, and we conclude that $P_t$
    is a \kl{strongly natural binomial polynomial} \kl{$\Nat$-rational
    polynomial}.

@@

\begin{definition}[Ultimately polynomial]
    \label{ultimately-polynomial:def}
    Let $\Sigma$ be a finite alphabet. 
    A function $f \colon \Sigma^* \to \Rel$
    is \intro{ultimately polynomial}
    when there exists $N_0 \in \Nat$ such that
    for all $n \in \Nat$,
    for all \kl{pumping pattern} $q \colon \Nat^n \to \Sigma^*$,
    there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
    such that:
    \begin{equation*}
        f \circ p = P
        \quad 
        \text{ over } (\Nat_{\geq N_0})^n
        \quad .
    \end{equation*}
\end{definition}
